\input{glyphtounicode}
\pdfgentounicode=1
\documentclass[runningheads]{llncs}

\usepackage{cmap}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{url}
\usepackage{color}
\usepackage{hyperref}
\usepackage{microtype}
\DisableLigatures{encoding = T1, family = *}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\urlstyle{rm}
\setlength{\emergencystretch}{1em}

\graphicspath{
  {../outputs/mobilenetv3_distilled_resnet50_full/}
  {../outputs/mobilenetv3_full/}
}

\title{Knowledge Distillation Across Teacher Architectures for
Edge-Efficient Plant Disease Classification}
\titlerunning{Distillation Across Teacher Architectures}

\author{
  Abdulbaki Celtik\inst{1} \and
  Meric Kelemeden\inst{1}
}

\authorrunning{A. Celtik and M. Kelemeden}

\institute{
  Department of Computer Engineering
}

\begin{document}
\maketitle

\begin{abstract}
Accurate image-based plant disease classification can support precision
agriculture, but high-capacity networks are not always convenient for edge or
mobile deployment. This work studies knowledge distillation from large teacher
networks into a compact MobileNetV3-Small student, and asks in particular
whether the \emph{architecture of the teacher} affects the distilled student.
As a preliminary demonstration, we distill a ResNet50 teacher into
MobileNetV3-Small on the 38-class PlantVillage dataset with a single
train/validation/test split; the distilled student reaches 0.9937 accuracy and
0.9907 macro-F1 with 1.56M parameters, about 15 times fewer than the teacher.
We then carry out the main study on the smaller and class-imbalanced Cassava
leaf-disease dataset (five classes) using stratified five-fold
cross-validation. Three teachers from different architectural families ---
ResNet50 (residual CNN), EfficientNet-B2 (compound-scaled CNN), and ViT-B/16
(vision transformer) --- are each retrained per fold and distilled into the
same student, and we report accuracy together with macro-averaged precision,
recall, and F1. On Cassava, distillation improves the student over a supervised
baseline by $1.4$--$1.8$ percentage points of accuracy, but the choice of
teacher has little effect: the three distilled students agree to within
$0.5$ points despite the teachers differing widely in accuracy
($82.4$--$85.1\%$) and in size ($7.7$M--$85.8$M parameters). Strikingly, the
weakest and largest teacher (ViT-B/16) yields a student on par with those of the
stronger convolutional teachers, and one that even exceeds its own teacher's
accuracy. These results suggest that, in this setting, the benefit of
distillation comes from the procedure itself rather than from the capacity or
architecture of the teacher.
\keywords{Knowledge distillation \and Plant disease classification \and
MobileNetV3 \and Cross-validation \and Vision transformer \and Edge inference.}
\end{abstract}

\section{Introduction}

Plant disease recognition is a practical computer vision problem with direct
relevance to crop monitoring and precision agriculture. Deep neural networks
can achieve strong classification accuracy on curated leaf image datasets, but
models with high parameter counts may be inconvenient for mobile or
edge-oriented deployment. This creates a tension between predictive accuracy
and computational efficiency.

This work studies that trade-off through knowledge distillation, in which a
high-capacity teacher network guides the training of a lightweight
MobileNetV3-Small student. The aim is not to claim deployment on a specific
device, but to evaluate whether a compact student can retain a teacher's
predictive behavior while using far fewer parameters. Beyond the standard
teacher-to-student comparison, we focus on a question that a single experiment
cannot answer: does the \emph{choice of teacher architecture} change how well
the student learns? To study this we compare three teachers from distinct
design families under a cross-validated protocol.

The contributions are as follows. First, as a preliminary demonstration, we
distill a ResNet50 teacher into MobileNetV3-Small on the 38-class PlantVillage
dataset using a single split. Second, and as the main study, we evaluate three
teacher architectures --- ResNet50, EfficientNet-B2, and ViT-B/16 --- distilled
into the same student on the Cassava leaf-disease dataset. Third, we use a
methodologically careful protocol: stratified five-fold cross-validation with
the teacher retrained from scratch on each fold's training partition, so that
the held-out fold never influences the teacher. Fourth, we report macro-averaged
precision and recall alongside accuracy and macro-F1, which is important given
the class imbalance of the Cassava data, present results as
mean~$\pm$~standard deviation across folds, and assess each distilled student
against the supervised baseline with a paired significance test.

\section{Background and Related Work}

PlantVillage provides a widely used collection of labeled plant leaf images for
plant health research \cite{hughes2015open}. The dataset has enabled controlled
studies of image-based plant disease classification, although its curated
imaging conditions are cleaner than many field scenarios. The Cassava
leaf-disease data used in the main study \cite{sankalana2021cassava} consists of
five classes (four disease categories and a healthy class) and is more
class-imbalanced, providing a harder and less controlled testbed. These two
datasets were chosen deliberately: PlantVillage is a clean, widely used
benchmark that is well suited to an initial controlled demonstration of the
distillation pipeline, whereas Cassava consists of real competition field
imagery with strong class imbalance, so it tests whether the same conclusions
hold under more realistic and challenging conditions. Both are also directly
relevant to the edge-deployment motivation, since plant disease diagnosis is a
natural use case for on-device inference in the field.

Deep learning has been applied extensively to plant disease classification.
Early work showed that convolutional networks trained on PlantVillage can
classify leaf diseases with high accuracy \cite{mohanty2016using}, and later
studies compared fine-tuning strategies across architectures for this task
\cite{too2019comparative}. Most of this literature focuses on maximizing
accuracy with large backbones; comparatively little attention has been paid to
compressing these models for on-device inference, which is the gap this work
addresses through distillation.

Knowledge distillation was introduced as a way to transfer information from a
larger teacher model to a smaller student model \cite{hinton2015distilling}.
Instead of learning only from one-hot labels, the student also learns from the
teacher's softened class probabilities, which can encode similarities between
classes, making distillation attractive when the deployment model must be
smaller than the model used to generate the strongest predictions. Beyond
matching output logits, later methods transfer intermediate representations,
such as hidden-layer hints \cite{romero2015fitnets} and spatial attention maps
\cite{zagoruyko2017attention}; a broad overview of these variants is given by
Gou et al.\ \cite{gou2021knowledge}. In this work we use the original
logit-based formulation, since our focus is the choice of teacher rather than
the distillation mechanism.

A recurring and directly relevant finding is that a more accurate teacher does
not necessarily produce a better student. Cho and Hariharan
\cite{cho2019efficacy} show that very large teachers can distill less
effectively into small students because of the capacity gap between them, and
Mirzadeh et al.\ \cite{mirzadeh2020improved} introduce intermediate
``teacher-assistant'' networks to bridge that gap. Beyer et al.\
\cite{beyer2022knowledge} further argue that the consistency of the
teacher-student training setup matters more than raw teacher accuracy. These
results motivate our central question: when the student is fixed, does the
teacher's \emph{architecture} --- and not merely its accuracy --- affect the
distilled student? Distillation has also been studied in the context of vision
transformers, most notably DeiT \cite{touvron2021training}, which distills
\emph{into} a transformer using a CNN teacher. Our setting is complementary: we
distill \emph{from} a transformer (and from two CNNs) into a compact
convolutional student, comparing teacher families under a single protocol.

The teachers in this work span three architectural families. ResNet50 is a
residual network that uses skip connections to ease optimization of deeper
convolutional models \cite{he2016deep}. EfficientNet-B2 is a convolutional
network obtained by compound scaling of depth, width, and resolution, and is
known for strong accuracy at a low parameter count \cite{tan2019efficientnet}.
ViT-B/16 is a vision transformer that treats an image as a sequence of patches
and relies on self-attention rather than convolution \cite{dosovitskiy2021image},
giving it a substantially different inductive bias from the two CNN teachers.
The student in all experiments is MobileNetV3-Small, an architecture designed
for mobile inference through efficient convolutional blocks and hardware-aware
design choices \cite{howard2019searching}. These models were chosen for a
specific reason. The three teachers are well-established, strong image
classifiers with pretrained ImageNet weights, but they represent deliberately
different design philosophies --- residual convolutions, compound-scaled
convolutions, and patch-based self-attention --- and they also differ widely in
parameter count, which lets us separate the effect of teacher architecture from
that of teacher size. MobileNetV3-Small was chosen as the single student
because it is a representative compact model intended for edge and mobile
deployment, making it the natural target for compression. Comparing CNN and
transformer teachers that distill into this same convolutional student lets us
probe whether teacher architecture, and not only teacher accuracy, influences
the student.

\section{Methodology}

Let $z_t$ and $z_s$ denote the teacher and student logits for an input image,
and let $y$ denote the ground-truth class label. Supervised teacher and
supervised student baselines are trained with cross-entropy against the hard
labels. For distillation, the teacher is first trained and then frozen, and the
student is optimized using a weighted sum of the cross-entropy loss and a
temperature-scaled Kullback--Leibler divergence term:

\begin{equation}
  \mathcal{L} =
  \alpha T^2
  \mathrm{KL}\left(
    \mathrm{softmax}\left(\frac{z_t}{T}\right)
    \,\middle\|\,
    \mathrm{softmax}\left(\frac{z_s}{T}\right)
  \right)
  + (1-\alpha)\mathrm{CE}(y, z_s).
  \label{eq:kd-loss}
\end{equation}

The $T^2$ factor follows the common practice of compensating for the gradient
scaling introduced by softened probabilities. The temperature is fixed at
$T=4.0$ throughout. The distillation weight is $\alpha=0.5$ in the preliminary
PlantVillage experiment and $\alpha=0.7$ in the main Cassava study; in both
cases $\alpha$ is held constant across teachers so that any difference between
teachers is attributable to the teacher rather than to the loss weighting.
MobileNetV3-Small is used as the student in every experiment because it has
substantially fewer parameters than any of the teachers while retaining a
convolutional design appropriate for image classification.

\section{Experimental Setup}

All experiments use pretrained ImageNet weights, images resized to
$224 \times 224$ pixels and normalized with ImageNet statistics, random
horizontal flipping and small random rotations during training, deterministic
resizing for validation and test, and the best validation checkpoint for final
evaluation. Inference latency is reported from a batch-one timing routine that
averages repeated forward passes; latency values are local hardware
measurements rather than hardware-independent deployment benchmarks.

\subsection{Main Study: Cross-Validated Multi-Teacher Distillation on Cassava}

The main study uses the Cassava leaf-disease dataset (five classes) and
stratified five-fold cross-validation. The dataset is partitioned into five
stratified folds, preserving class proportions. For each fold, that fold serves
as the test set, and the remaining data is split (again with stratification)
into training and validation. Crucially, the teacher is \emph{retrained from
scratch on that fold's training partition} before being frozen and used to
distill the student. This avoids the optimistic bias that would arise if a
single teacher had been exposed to images that later appear in a fold's test
set. The fold partition is fixed by seed, so every teacher and every model sees
identical folds. All models are initialized from ImageNet-pretrained weights, so
the comparison concerns transfer to the plant-disease task rather than training
from random initialization; this pretraining is shared across all teachers and
the student and therefore does not bias the teacher comparison.

Three teachers are evaluated --- ResNet50, EfficientNet-B2, and ViT-B/16 ---
each distilled into a MobileNetV3-Small student, and compared against a
supervised MobileNetV3-Small baseline. The student schedule and distillation
hyperparameters ($T=4.0$, $\alpha=0.7$) are held constant across teachers.
Per-fold teacher schedules and the student schedule are summarized in
Table~\ref{tab:cv-config}. Each teacher is also evaluated on its own, trained
supervised under the same cross-validation, so that the teachers' own
cross-validated performance can be reported alongside the students. These runs
use an NVIDIA RTX 4070 (CUDA) with automatic mixed precision. For each model we
report the mean and standard deviation across the five folds of accuracy and of
macro-averaged precision, recall, and F1; macro averaging weights each class
equally, which matters because the Cassava classes are imbalanced. Because the
folds are shared across models, each distilled student can be compared against
the supervised baseline with a paired two-sided $t$-test on the five per-fold
accuracies; we report the resulting $p$-values so that differences are not
over-interpreted given the small number of folds.

\begin{table}
  \centering
  \caption{Per-fold training configuration for the Cassava cross-validation
  study. All teachers are distilled into the same MobileNetV3-Small student
  with $T=4.0$ and $\alpha=0.7$.}
  \label{tab:cv-config}
  \begin{tabular}{lllll}
    \toprule
    Model & Role & Learning rate & Epochs & Batch \\
    \midrule
    ResNet50 & Teacher & $3\times 10^{-4}$ & 10 & 64 \\
    EfficientNet-B2 & Teacher & $3\times 10^{-4}$ & 10 & 64 \\
    ViT-B/16 & Teacher & $1\times 10^{-4}$ & 10 & 32 \\
    MobileNetV3-Small & Student / baseline & $5\times 10^{-4}$ & 15 & 64 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Preliminary Study: Single-Split Distillation on PlantVillage}

As a preliminary demonstration, a ResNet50 teacher, a supervised
MobileNetV3-Small baseline, and a distilled MobileNetV3-Small student are
trained on the full 38-class PlantVillage dataset with a single deterministic
70/15/15 train/validation/test split (seed 42). These runs use 15 epochs,
weight decay $10^{-4}$, batch size 32, and distillation settings $T=4.0$ and
$\alpha=0.5$, and were executed on Apple Silicon (MPS). The ResNet50 teacher
uses learning rate $3\times 10^{-4}$, the supervised student uses $10^{-3}$,
and the distilled student uses $5\times 10^{-4}$.

\section{Results and Discussion}

\subsection{Main Study: Cassava Cross-Validation}

Table~\ref{tab:cv-results} summarizes the cross-validated results on Cassava.
The table reports each teacher trained supervised, the supervised
MobileNetV3-Small baseline (teacher ``--''), and the student distilled from
each teacher, as mean~$\pm$~standard deviation across the five folds. Two
questions are of interest: whether distillation improves the student over the
supervised baseline on this harder, imbalanced dataset, and whether a stronger
or architecturally different teacher distills better than ResNet50.

\input{../outputs/cv_summary.tex}
% Regenerated by `python -m src.collate_cv` after the cross-validation runs;
% until then it shows placeholder dashes.

Distillation helps. All three teachers raise the student above the supervised
baseline (81.67\% accuracy) by a similar margin: $+1.82$ points for ResNet50
(83.49\%), $+1.45$ for ViT-B/16 (83.12\%), and $+1.37$ for EfficientNet-B2
(83.04\%). In the paired $t$-test the ResNet50 and ViT-B/16 gains are
statistically significant ($p=0.029$ and $p=0.036$), whereas the
EfficientNet-B2 gain does not reach significance ($p=0.079$); given the
consistent positive mean difference, this most likely reflects the limited power
of five folds together with that run's larger fold-to-fold variance rather than
an absence of effect. The macro-F1 scores improve in step with accuracy, so the
gains are not confined to the majority class.

The more notable result is how little the teacher matters. The three distilled
students lie within $0.5$ points of one another (83.04--83.49\%), a spread no
larger than their across-fold standard deviations, even though the teachers
differ substantially in cross-validated accuracy (82.40--85.13\%), in
architectural family (two CNNs and one transformer), and in size (7.7M, 23.5M,
and 85.8M parameters). Teacher accuracy is therefore a poor predictor of student
accuracy here: the weakest and by far the largest teacher, ViT-B/16, produces a
student as good as the one distilled from the strongest teacher, and does so
with a statistically significant improvement over the baseline. The ViT-B/16 student (83.12\%) in
fact surpasses its own teacher (82.40\%), which is inconsistent with the view
that the student simply inherits the teacher's accuracy. It also runs counter to
the expectation that a very large teacher, separated from a 1.5M-parameter
student by a wide capacity gap, would distill poorly. Taken together, these
observations indicate that in this setting the improvement comes from the
distillation procedure --- the softened-label regularization --- more than from
the specific teacher.

This has a practical implication for the edge-deployment motivation. Because the
deployed model is always the same 1.5M-parameter student and the teacher is used
only during training, a smaller and cheaper teacher is sufficient:
EfficientNet-B2, at roughly a third of ResNet50's parameters, teaches about as
well, and the much larger ViT-B/16 offers no advantage as a teacher despite its
cost. The deployment cost is unchanged by the teacher in any case: all four
students share the MobileNetV3-Small architecture and therefore have essentially
the same batch-one inference latency (approximately 5\,ms on the RTX 4070), and
the teacher is discarded after training. Reporting
mean~$\pm$~standard deviation across folds, rather than a single split, makes
these comparisons more robust to the particular partition of the data.

\subsection{Sensitivity to Distillation Hyperparameters}

The main study fixes the distillation temperature and weight at the widely used
values $T=4.0$ and $\alpha=0.7$, and holds them constant across teachers so that
differences are attributable to the teacher. To check that this choice is
reasonable and that the comparison is not an artifact of a particular setting,
we run a single-split sensitivity analysis for the ResNet50 teacher on fold~0 of
the Cassava partition. Because $T$ and $\alpha$ affect only the student loss, the
teacher is trained once and reused for every setting; we then sweep
$T \in \{1, 2, 4, 8\}$ at $\alpha=0.7$ and $\alpha \in \{0.3, 0.5, 0.7, 0.9\}$ at
$T=4.0$. Table~\ref{tab:ablation} reports the resulting student accuracy and
macro-F1.

\input{../outputs/ablation_resnet50_cassava/ablation_summary.tex}
% Regenerated by `python -m src.ablate_kd --config configs/ablation_resnet50_cassava.yaml`.

The student is stable across both sweeps: accuracy stays within roughly one
point over all settings (82.78--84.14\%), with a gentle maximum exactly at the
default $T=4.0$, $\alpha=0.7$ used in the main study (84.14\%). Very low
temperatures, very light teacher weighting ($\alpha=0.3$), and very heavy
teacher weighting ($\alpha=0.9$) are all slightly worse, and $\alpha=0.9$ in
particular lowers macro-F1 (to 69.39\%), consistent with over-weighting the
teacher at the expense of the minority classes. The flat response confirms that
the fixed configuration used in the main comparison is well justified and that
the teacher comparison is not an artifact of a particular hyperparameter choice.
These results are single-split point estimates without fold error bars, so they
are intended as a sensitivity check rather than as a second cross-validated
comparison.

\subsection{Preliminary Study: PlantVillage Single Split}

Table~\ref{tab:results} reports the preliminary single-split results on the
38-class PlantVillage task. The ResNet50 teacher reaches 0.9936 test accuracy
and 0.9912 macro-F1. The supervised MobileNetV3-Small baseline is already
strong, with 0.9851 accuracy and 0.9779 macro-F1. After distillation from
ResNet50, MobileNetV3-Small improves to 0.9937 accuracy and 0.9907 macro-F1
while retaining the same 1.56M parameter count.

\begin{table}
  \centering
  \caption{Preliminary single-split test results on the 38-class PlantVillage
  task.}
  \label{tab:results}
  \resizebox{\textwidth}{!}{
  \begin{tabular}{llllrr}
    \toprule
    Model & Training strategy & Accuracy & Macro-F1 & Parameters & Latency \\
    \midrule
    ResNet50 & Teacher supervised & 0.9936 & 0.9912 & 23,585,894 & 5.47 ms \\
    MobileNetV3-Small & Supervised baseline & 0.9851 & 0.9779 & 1,556,806 & 3.51 ms \\
    MobileNetV3-Small & Distilled from ResNet50 & 0.9937 & 0.9907 & 1,556,806 & 3.53 ms \\
    \bottomrule
  \end{tabular}}
\end{table}

The comparison between the supervised and distilled MobileNetV3-Small models is
informative because the architecture and parameter count are identical in both
rows. The improvement from 0.9851 to 0.9937 accuracy therefore does not come
from increasing student capacity; it indicates that the teacher logits provide
useful training signal beyond the hard labels alone. At the same time,
PlantVillage images are relatively clean and the supervised student is already
very strong, which limits the headroom for distillation. This motivates the
main cross-validated study on the harder Cassava data, where the supervised
baseline is weaker and the effect of the teacher can be observed more clearly.

\section{Reproducibility and Limitations}

All experiments are configuration-driven. The preliminary PlantVillage runs
correspond to the full-dataset teacher, student, and distillation
configurations, while the main study is driven by the cross-validation configs
and executed with the \texttt{src.cross\_validate} runner; results are collated
into a single table with \texttt{src.collate\_cv}. Reported metrics are taken
from the stored result files rather than copied from terminal logs, so the
report stays tied to reproducible artifacts: each run stores a model summary,
per-fold metrics, final aggregated metrics, and result CSVs.

Several limitations remain. First, the distillation hyperparameters are kept
fixed across teachers ($T=4.0$, with $\alpha=0.5$ or $0.7$); a more complete
study would tune temperature and loss weighting per teacher, which would help
separate the effect of the teacher architecture from that of the distillation
schedule. Second, the conclusion that the teacher matters little rests on the distilled
students differing by no more than their across-fold variation; with only five
folds the statistical power is limited, so this is evidence that any teacher
effect is small rather than proof that it is exactly zero, and more folds or
repeated runs would tighten the comparison. Third, the per-fold teacher
retraining means each teacher is trained multiple times, which is
computationally expensive but necessary to avoid test leakage. Fourth, latency is measured locally --- on Apple Silicon (MPS) for the
preliminary study and on an NVIDIA RTX 4070 (CUDA) for the main study --- so
latency values are comparable only within each study and should not be read as
deployment benchmarks. Finally, both datasets are curated; performance on field
images with cluttered backgrounds, varying lighting, and different cameras may
be lower.

\section{Conclusion}

This work studied knowledge distillation from large teachers into a compact
MobileNetV3-Small student for edge-efficient plant disease classification. A
preliminary single-split experiment on the 38-class PlantVillage dataset showed
that a ResNet50 teacher can be distilled into a student with 0.9937 accuracy and
0.9907 macro-F1 at about 15 times fewer parameters than the teacher. The main
study then used stratified five-fold cross-validation on the harder, imbalanced
Cassava dataset to compare three teacher architectures --- ResNet50,
EfficientNet-B2, and ViT-B/16 --- distilled into the same student, reporting
accuracy together with macro-averaged precision, recall, and F1. On Cassava,
distillation raised the student by $1.4$--$1.8$ accuracy points over the
supervised baseline, a gain that was statistically significant for the ResNet50
and ViT-B/16 teachers. The choice of teacher, however, had little effect: the
three distilled students agreed to within half a point despite teachers that
differed widely in accuracy, architecture, and size, and the weakest and largest
teacher (ViT-B/16) taught as effectively as the strongest while its student even
surpassed it. The benefit of distillation in this setting therefore appears to
stem from the procedure rather than from the capacity or architecture of the
teacher, which is encouraging for edge deployment because a small, inexpensive
teacher suffices. The main limitations are dataset realism and the use of fixed
distillation hyperparameters; future work should tune the distillation schedule
per teacher, evaluate on more realistic field imagery, and measure latency on
actual edge hardware.

\bibliographystyle{splncs04}
\bibliography{references}

\end{document}
